Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-08-23 papers/paper/ms.tex
update ms

% upon AAS submission
%\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
%\documentclass[11pt,twocolumn,tighten]{aastex63}
\turnoffedit

@@

{\it Crowding (bit 8)}---We searched the Gaia DR3 point source catalog
for stars within 1 Kepler pixel (4$''$) of every target star.  While
such companions may not not be physically associated with the target
star, their presence can confuse rotation period measurements.  We
therefore flagged any stars with neighbors down to 1/10$^{\rm th}$ the
brightness of the target star within this region ($\Delta G < 2.5$).

@@

Figure~\ref{fig:gyroage_vs_teff} is a visualization of $t_{\rm gyro}$
vs.~$T_{\rm eff}$ for our rotation-based age catalog. 
Each bar denotes the $\pm$1$\sigma$ uncertainty for a star's
rotation-based age:
this plot can be viewed as the transformation of
the left panel of Figure~\ref{fig:prot_vs_teff}.
The overdensity of $P_{\rm rot}$$\approx$20\,day G dwarfs corresponds
to the large overdensity at $t_{\rm gyro}$$\approx$3\,Gyr.
The (non-physical) deficit of $\sim$22\,day rotators appears as a
deficit between 5500--6000\,K.
A second deficit is also visible from 3800--5000\,K;
it is associated with the ``intermediate period gap'', which may be
associated with a transition from spot-dominated to faculae-dominated
light curves \citep[e.g.][]{Reinhold2019}.
At $\approx$1\,Gyr, stalled spin-down yields larger uncertainties for
K dwarfs.
The pile-up near $\approx$5\,Gyr is imposed by our choice of prior;
given that \texttt{gyro-interp} returns non-calibrated ages in this
regime, we truncated our analysis at 4\,Gyr.

@@

This analysis is particularly interesting in relation to the question
of how fast stars $\gtrsim$4\,Gyr old spin down.  If we lived in a
universe in which stellar rotation rates and spot-induced photometric
amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
stars \replaced{would tend to }{could }contaminate the 2.5-3\,Gyr peaks in
Figures~\ref{fig:hist_tgyro} and~\ref{fig:hist_tgyro_mcq}.  This
scenario deserves consideration, given\replaced{ evidence from asteroseismic
rotation periods that has pointed to}{ {\it i)} asteroseismic evidence for} slowed spin-down rates after
$\approx$4\,Gyr for Sun-like stars
\citep[e.g.][]{vanSaders_2016,2024ApJ...962..138S}\added{, and {\it ii)} the observation of a ``long-period edge''
in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
may pile up over similar timescales \citep{David_2022}}.  However, \replaced{because
the \citeauthor{McQuillan_2014} sample does not include many stars
older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this analysis}{the analysis
by \citet{2022ApJ...937...94M} fitted isochrones to a high-quality subset of Kepler data,
and found that the
sample of detected rotators from \citeauthor{McQuillan_2014} does not include many
stars older than $\approx$\,4\,Gyr.}
\replaced{would argue}{This suggests} that the 2.5-3\,Gyr peak \replaced{cannot}{is unlikely to} be caused by a pile-up of
much older stars with especially lethargic spin-down rates.

>>>>

% upon AAS submission
\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
%\documentclass[11pt,twocolumn,tighten]{aastex63}
\turnoffedit

@@

{\it Crowding (bit 8)}---We searched the Gaia DR3 point source catalog
for stars within 1 Kepler pixel (4$''$) of every target star.  While
such companions may not not be physically associated with the target
star, their presence can confuse rotation period measurements.  We
therefore flagged any stars with neighbors down to 1/10$^{\rm th}$ the
brightness of the target star within this region ($\Delta G$$<$2.5).\added{ We also
considered a deeper cut ($\Delta G$$<$5), and found it had negligible impact
on our conclusions.}

@@

Figure~\ref{fig:gyroage_vs_teff} is a visualization of $t_{\rm gyro}$
vs.~$T_{\rm eff}$ for our rotation-based age catalog. 
Each bar denotes the $\pm$1$\sigma$ uncertainty for a star's
rotation-based age:
this plot can be viewed as the transformation of
the left panel of Figure~\ref{fig:prot_vs_teff}.
The overdensity of $P_{\rm rot}$$\approx$20\,day G dwarfs corresponds
to the large overdensity at $t_{\rm gyro}$$\approx$3\,Gyr.
The (non-physical) deficit of $\sim$22\,day rotators appears as a
deficit between 5500--6000\,K.
A second deficit is also visible from 3800--5000\,K;
it is associated with the ``intermediate period gap'', which may be
associated with a transition from spot-dominated to faculae-dominated
light curves \citep[e.g.][]{Reinhold2019}\added{, or with rapid spin-down at Rossby numbers near 0.5 \citep{Lu2022}}.
At $\approx$1\,Gyr, stalled spin-down yields larger uncertainties for
K dwarfs.
The pile-up near $\approx$5\,Gyr is imposed by our choice of prior;
given that \texttt{gyro-interp} returns non-calibrated ages in this
regime, we truncated our analysis at 4\,Gyr.

@@

This analysis is particularly interesting in relation to the question
of how fast stars $\gtrsim$4\,Gyr old spin down.  If we lived in a
universe in which stellar rotation rates and spot-induced photometric
amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
stars \replaced{would tend to }{could }contaminate the 2.3-3\,Gyr peaks in
Figures~\ref{fig:hist_tgyro} and~\ref{fig:hist_tgyro_mcq}.  This
scenario deserves consideration, given\replaced{ evidence from asteroseismic
rotation periods that has pointed to}{ {\it i)} asteroseismic evidence for} slowed spin-down rates after
$\approx$4\,Gyr for Sun-like stars
\citep[e.g.][]{vanSaders_2016,2024ApJ...962..138S}\added{, and {\it ii)} the observation of a ``long-period edge''
in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
may pile up over similar timescales \citep{David_2022}}.  However, \replaced{because
the \citeauthor{McQuillan_2014} sample does not include many stars
older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this
analysis}{this pileup is thought to happen at $>$5\,Gyr for
late-G dwarfs and early K dwarfs \citep[see Figure~4
by][]{2024ApJ...962..138S}.  This means that it cannot affect the
2.3\,Gyr peak in the age distribution that we see for 4400-5400\,K stars.   
Moreover, while some contamination by older stars may be present in the peaks,
the analysis by \citet{2022ApJ...937...94M} fitted
isochrones to a high-quality subset of Kepler data, and found that the
\citeauthor{McQuillan_2014} sample does not
include many stars older than $\approx$\,4\,Gyr.}
\replaced{would argue}{These lines of evidence suggest} that the peak
\replaced{cannot be}{in the age distribution is not} caused by a pile-up of
much older stars with especially lethargic spin-down rates\added{, but that it could be
contaminated by them, especially for stars hotter than the Sun}.
